\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=violet,urlcolor=violet,citecolor=violet]{hyperref}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.1ex}{0.6ex}
\titlespacing*{\subsection}{0pt}{0.8ex}{0.4ex}
\setlength{\parskip}{0.35em}
\setlength{\parindent}{0pt}

\title{\textbf{Tenet: Agent Memory as a Self-Consistent Belief State}}
\author{Anas\\\small Global AI Hackathon with Qwen Cloud (Track 1), 2026\\
\small\texttt{https://github.com/Nas01010101/tenet}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
\noindent Long-term memory for LLM agents is almost universally implemented as
\emph{retrieval over a growing log of past turns}---a document-retrieval abstraction. We
argue this is the wrong abstraction for an agent that must model a changing world. We
introduce \textbf{knowledge churn}---the repeated updating of a fact over a long
interaction---and show that retrieval-augmented memory (RAG-memory) \emph{silently
degrades} under it: as the number of stale versions of a fact exceeds the retrieval budget
$k$, the reader is handed conflicting values and answers incorrectly. On a controlled
benchmark, a strong RAG-memory falls from $100\%$ to $50\%$ current-value accuracy as a
fact is updated $2\!\to\!12$ times. We propose \textbf{Tenet}, which reframes memory as a
\textbf{self-consistent belief state}---a compact, supersession-aware set of facts about the
user---rather than a document store. Tenet (i)~distills raw turns into atomic, keyed facts; (ii)~maintains
a \textbf{bi-temporal} record so a changed fact \emph{supersedes} its predecessor (retired
to history, not overwritten); (iii)~enforces \textbf{belief--evidence consistency} by
retiring raw evidence that echoes a superseded belief; (iv)~applies a
\textbf{predictive-coding write policy}---surprise-gating---that stores only observations the
model cannot already predict; and (v)~closes the accuracy gap to raw retrieval with
\textbf{belief-anchored evidence expansion}---spending spare context on query-relevant turns
from the sessions the belief state already surfaced. Tenet holds \textbf{$100\%$ current-value
accuracy across all levels of the templated single-attribute churn primitive} (where a strong
RAG-memory falls to $50\%$; on a harder \emph{paraphrased} multi-attribute variant this claim
is falsified and recovered by a read-time consistency fix---full preprint / \texttt{BENCHMARK.md}~\S9),
matches strong
RAG on retrieval recall ($95$--$97.5\%$), and---with expansion---\textbf{matches its one-shot
answer accuracy at equal-or-lower token budget} ($57.5\%$ vs $57.5\%$, gpt-4o reader) while
retaining a high-efficiency point at \textbf{half the context} and the \textbf{best
accuracy-per-token} of the systems we evaluate. Tenet thus traces an accuracy--efficiency
\emph{frontier} that meets RAG at its budget and beats it at every lower one. We release all code.
\end{abstract}

\section{Introduction}
An agent that talks to a user for months does not need a transcript; it needs a
\emph{model of the user} that stays true as the user changes. Yet the dominant memory
design---retrieval-augmented generation over stored turns~\cite{mem0,longmemeval}---treats
memory as a document index: embed every turn, retrieve the top-$k$ most similar, let the
reader sort it out. This works for one-shot recall of a \emph{static} fact. It fails,
quietly, when facts \textbf{change}.

We formalize this as \textbf{knowledge churn}. A user who moves cities several times
generates many ``I moved to $X$'' turns; all are similar to ``where do I live?'', so the
top-$k$ fills with \emph{stale} versions. Once the number of updates exceeds $k$, the latest
value may not be retrieved at all, and even when it is, the reader must infer recency from
contradictory statements. Accuracy collapses (\S\ref{sec:churn}).

The root problem is abstraction. A document store has no notion that ``I live in Boston''
and ``I live in Seattle'' are the \emph{same fact} with a \emph{changed value}. We argue
memory should be a \textbf{belief state}: a compact set of \emph{current} beliefs, each with
a temporal extent, updated by observation, kept internally consistent, and queryable across
time---the stance predictive-coding accounts take toward
perception~\cite{friston}, brought to agent memory.

\textbf{Contributions.}
\begin{enumerate}\itemsep2pt
\item We identify and name \textbf{knowledge churn}, a failure mode of retrieval memory, and
give a controlled benchmark exhibiting it (RAG: $100\%\!\to\!50\%$; \S\ref{sec:churn}).
\item We present \textbf{Tenet}, a memory that is a self-consistent belief state:
bi-temporal supersession, a \textbf{belief--evidence consistency} rule, and a
\textbf{surprise-gated} (predictive-coding) write policy.
\item We evaluate on LongMemEval\_S and controlled tests: Tenet is \textbf{churn-robust
($100\%$)}, on par with RAG on recall ($95\%$), \textbf{best on accuracy-per-token}, and---with
belief-anchored evidence expansion---\textbf{at parity with strong RAG on one-shot accuracy at
equal token budget}, closing a gap belief-only compression left open.
\item On the standardized \textbf{MemoryAgentBench FactConsolidation} benchmark,
ingestion-time supersession with zero-LLM deterministic keys \textbf{exceeds the published
single-hop SOTA at the gpt-4o-mini tier (86.5 vs 78.0 pooled) and ties the multi-hop tier
(30.2)}---using only a local 7B backbone, where the benchmark's 22 systems score
$\le$60~/~$\le$7.
\end{enumerate}

\section{Related Work}
\textbf{Retrieval memory.} Mem0~\cite{mem0} distills salient facts at write time over a
vector store; it attaches only a \emph{creation} timestamp, and its own report prices the graph
variant at ${\sim}1.8\times$ p95 latency for a ${\sim}2\%$ quality gain---evidence we
weigh in choosing a light substrate. LongMemEval~\cite{longmemeval} is the standard
long-horizon benchmark; its V2 adds a \emph{latency-aware} metric, a field shift toward
accuracy \emph{per cost} that our per-token results target.
\textbf{Temporal knowledge graphs.} Zep/Graphiti~\cite{zep} maintain a bi-temporal graph
with automatic invalidation---but pay heavy per-write extraction
and require graph infrastructure; Tenet keeps the bi-temporal semantics without the graph.
\textbf{The 2026 bi-temporal convergence.} Concurrent systems adopted bi-temporal
supersession: MemStrata~\cite{memstrata} applies deterministic $(s,r,o)$ supersession over
a bi-temporal ledger with no LLM on the read path---and shows \emph{similarity-threshold}
supersession leaks stale values where deterministic keying does not, independently
corroborating our keyed design; Engram~\cite{engram} pairs a bi-temporal graph with a
hybrid facts-plus-chunks read path (converging on our dual-pool finding);
TOKI~\cite{toki} formalizes contradiction resolution as a bitemporal operator algebra;
and \cite{freshness} sets the current MemoryAgentBench FactConsolidation SOTA with
deterministic post-retrieval aggregation. Tenet differs in \emph{where} consistency is
enforced---at ingestion and at recall (belief--evidence consistency)---and in coupling the
belief state to surprise-gated writes and budget-bounded evidence expansion, evaluated on
the knowledge-churn axis none of them report.
\textbf{Memory in the Qwen ecosystem.} Alibaba's own 2026 memory line takes the opposite
design point: QwenLong-L1.5~\cite{qwenlong} reaches 76.4 on LongMemEval by RL-training a
30B reasoner that reads up to 128K tokens per query; AgeMem~\cite{agemem} RL-trains the
memory policy itself (Qwen2.5-7B backbone); ActMem~\cite{actmem} (Alibaba-co-authored)
reaches 75.6 with LLM-built causal graphs at ingestion. All three spend heavily---at
training time, ingestion time, or read time. Tenet reaches 60.0 at the same reader tier
with \textbf{zero-LLM ingestion and $\sim$2K read tokens per query} ($\sim$79\% of
QwenLong-L1.5's score at under 2\% of its read budget), and its structural wins survive
on a 7B backbone.
\textbf{Active memory navigation.} A concurrent 2026 line makes memory access itself
agentic. NapMem~\cite{napmem} exposes a four-level memory pyramid as a structured action
space and trains a GRPO policy that descends it and stops once evidence is sufficient
(trained 9B beats untrained 397B on memory benchmarks); QwenLong-L1.5~\cite{qwenlong}
plans a navigational path over memory chunks (LongMemEval $+15.6$); MemFlow
\cite{memflow} is training-free but LLM-mediated (router + LLM-Validator retry);
MemReader~\cite{memreader} moves active navigation to the write side; AgeMem
\cite{agemem} and Mem-$\alpha$~\cite{memalpha} RL-train the policy itself; MemCog
\cite{memcog} reports a similar frame (affiliation unverified). All make the \emph{read}
path agentic; Tenet keeps reads LLM-free and gets adaptive depth from an embedding
saturation gate (\S\ref{sec:navigate}) instead of a learned stop.
\textbf{OS-style / observational memory.} MemGPT/Letta~\cite{memgpt} page memory between
context and archival storage; observational schemes keep a stable summary. Both are largely
append-oriented and do not treat supersession or forgetting as first-class.
\textbf{What is missing.} No prior system combines bi-temporal supersession,
\emph{belief--evidence consistency}, \emph{predictive-coding write-gating}, principled
forgetting, and \emph{belief-anchored evidence expansion} in a light, graph-free
store---nor reports the knowledge-churn regime.

\section{Method}
Tenet stores two layers over one bi-temporal table: a \textbf{belief layer} of distilled,
keyed facts, and an \textbf{evidence layer} of raw turns. Reads never call an LLM.

\textbf{Distillation.} Each turn is distilled into atomic facts, each with a stable
semantic key $\kappa=\textit{subject}\!:\!\textit{attribute}$ (e.g.\ \texttt{user:residence}),
a salience $s\in[0,1]$, and an event time. The key makes supersession reliable: embedding
similarity cannot separate a \emph{restated} fact from a \emph{value-changed} one (we measure
a value change at cosine $0.99$---\emph{higher} than a plain rephrasing of the same fact at
$0.79$, so no similarity threshold separates them), but a shared key can.

\textbf{Bi-temporal supersession.} Every memory carries event time
($\texttt{valid\_at},\texttt{invalid\_at}$) and transaction time
($\texttt{created\_at},\texttt{expired\_at}$). Storing a fact at key $\kappa$ whose value
differs from the current one \emph{supersedes} it: the old fact's
$\texttt{invalid\_at}/\texttt{expired\_at}$ are set; it leaves the current set but stays in
history. Current recall filters $\texttt{expired\_at}=\varnothing$; $\text{recall}(as\_of{=}t)$
reconstructs the belief set at any past $t$ (time-travel).

\textbf{Belief--evidence consistency (key rule).} The evidence layer answers detail
questions but is where stale values hide: ``I moved to Boston'' survives after the belief
moves on. We retire from current recall any raw slice $e$ close to a \emph{superseded} belief:
\[
\text{exclude } e \iff \max_{f\in\mathcal{F}_{\text{expired}}}\cos(e,f)\ \ge\ \tau_{\text{stale}},
\qquad \tau_{\text{stale}}=0.80.
\]
This single rule turns supersession into end-to-end correctness: $55\%\!\to\!100\%$
(\S\ref{sec:abl}).

\textbf{Predictive-coding write policy.} Predictive coding stores \emph{prediction error},
not everything. On write, a raw observation $e$ is discarded if the store already predicts it:
\[
\text{store } e \iff \max_{e'\in\text{store}}\cos(e,e')\ <\ g_{\text{surprise}},
\qquad g_{\text{surprise}}=0.97 .
\]
This bounds the store and drops redundant repetition with no accuracy loss (\S\ref{sec:abl}).

\textbf{Forgetting.} Rank $=$ relevance $\times$ decay, with
$d = 2^{-\Delta t/h}\,(1+\beta\log(1{+}\text{uses}))\,(0.6+0.8s)$, half-life $h=14$\,d; a
sweep archives low-$d$ unpinned memories. Retrieval is a \emph{dual pool} (beliefs for
consistency, evidence for verbatim detail), each guaranteed a budget share.

\textbf{Fact dynamics --- learning how facts change.} The ledger doubles as training
data for drift: superseded facts are observed lifetimes, current facts censored ones.
Per key class a conjugate Gamma--exponential model gives closed-form survival
$P(\text{valid}\mid\Delta t)=\big(\tfrac{\beta_0+T}{\beta_0+T+\Delta t}\big)^{\alpha_0+n}$
--- ``residence'' learns a slow hazard, ``mood'' a fast one, per user, no LLM.
Co-supersession statistics add \emph{ripple}: a correlated key's change discounts its
neighbors' survival, so one observation updates beliefs about unobserved attributes.
Surfaced as per-fact \texttt{confidence} + an \texttt{uncertain\_facts()}
re-verification list; deliberately annotation-only (rank-demotion measurably
re-created the churn failure before we reverted it). A trained alternative---a
276k-param GRU temporal point process (Weibull hazard + next-key + contrastive
next-value heads)---beats the closed form decisively on planted non-memoryless
structure (NLL $2.76\!\to\!1.99$, CI excludes 0; next-value 0.997 recall@5 vs 0.05
chance; 5 seeds) but is honestly mixed on sparse real chains; it ships as a 1\,MB
numpy artifact ($106\,\mu$s/query), opt-in, defaults unchanged.

\textbf{Belief-anchored evidence expansion.} Compressing a session into a few beliefs wins
churn and efficiency but can drop the detail a multi-hop question needs, even when the right
session \emph{is} retrieved (recall $95$--$100\%$). The top-$k$ result names both the belief
state \emph{and the sessions it came from} (each memory's \texttt{source}). Given spare
budget $B$, Tenet fills it with query-relevant raw turns drawn \emph{only from those
already-surfaced sessions}---evidence anchored to the belief state, not the whole
haystack---under the same stale-echo filter. Setting $B$ to the baseline RAG budget makes
expansion never spend more context than flat retrieval; the count $m$ is a knob ($m{=}0\to$
efficiency point; $m$ large under $B\to$ parity point) that lets one system trace a frontier.

\textbf{Saturation-gated navigation.}\label{sec:navigate} Associative multi-hop recall
re-scores the store at a caller-fixed depth $N$; \texttt{navigate()} instead deepens
iteratively, reusing belief-anchored expansion at each hop, and adopts hop $d{+}1$ only
while its new evidence stays informative:
\[
\text{adopt hop } d{+}1 \iff \max_{e\in\textit{new}_{d+1}}\textit{score}(e) \ge
\tau_{\text{gain}} = 0.15,
\]
stopping (budget $\le4$ hops) once no hop clears the gate. The gate is
embedding-only---no LLM, no training---the training-free counterpart to NapMem's
learned stop and MemFlow's LLM-Validator retry, at $\sim$ms latency. Validated as a
\emph{mechanism} on a deterministic synthetic store (bridge fact reached via an
associative hop; early-stop at hop 2 on a saturated query vs.\ hop 3 on a multi-hop
one). A subsequent seeded A/B at the \texttt{qwen3.7-plus} tier (FC 6k cells,
$n{=}20$) measured \emph{no benefit}: identical accuracy to default recall, $-5$
to $-10$pp vs.\ a fixed-4-hop arm, all within Wilson CIs---a strong reader is
robust to the larger fixed pool, so adaptive early-stopping buys no precision.
We report navigation as a mechanism, not a measured gain.

\section{Experiments}
\textbf{Protocol.} LongMemEval\_S (500 questions, ${\sim}115$k-token histories).
We use a \textbf{\texttt{gpt-4o} reader} (the same reader Mem0 and Zep report against), a local embedder (\texttt{bge-small}), and a cheap \texttt{gpt-4o-mini} distiller; the shipped system runs the same code on Qwen Cloud (\texttt{text-embedding-v4}, \texttt{qwen3.7-plus}) by a config flip. Numbers are compared only to baselines run under identical settings.
Baselines: \textbf{RAG} (top-$k$ raw turns) and \textbf{full-context} (entire history).

\subsection{Recall, and the accuracy--efficiency frontier (n$=$40, gpt-4o reader)}
\label{sec:frontier}
\begin{center}\small
\begin{tabular}{llrrrr}
\toprule
System & mode & recall@10 & QA acc & reader tok & \textbf{acc/1k tok} \\
\midrule
full-context & --- & --- & 65\%$^{*}$ & ${\sim}124\text{k}$ & 0.5$^{*}$ \\
RAG          & top-$k$ turns & 95\% & 57.5\% & 2{,}101 & 27.4 \\
\textbf{Tenet} & efficiency ($m{=}0$) & \textbf{97.5\%} & 52.5\% & \textbf{1{,}067} & \textbf{49.2} \\
\textbf{Tenet} & parity (expansion) & \textbf{97.5\%} & \textbf{57.5\%} & 2{,}083 & 27.6 \\
\bottomrule
\end{tabular}
\end{center}
Tenet is a \emph{frontier}, not a point. At its \textbf{efficiency} point it answers at half
the context ($1{,}067$ tok) for the best accuracy-per-token of any system ($49.2$, $1.6\times$
RAG) at recall parity. Turning up belief-anchored expansion under RAG's own budget brings raw
QA accuracy to \textbf{parity with a strong RAG ($57.5\%{=}57.5\%$) at fewer tokens}, closing
the one-shot gap belief-only compression left open. Tenet thus meets RAG's accuracy at its
budget and beats it at every lower one; the one category still behind is \emph{multi-session}
synthesis ($42.9$ vs $57.1$, up from $28.6$), where a question needs several evidence sessions
but only some are surfaced. $^{*}$Full-context, at a weaker reader, spends $100\times$ the
tokens for no gain over RAG. Reader-robust: on a cheaper \texttt{gpt-4o-mini} reader the parity
point edges ahead (Tenet $60.0$ vs RAG $55.0$); per-token dominance holds across the
\texttt{gpt-4o-mini} and \texttt{gpt-4o} readers we ran ($\approx$$1.6\times$).

\subsection{Knowledge churn (headline)}
\label{sec:churn}
One fact updated $N$ times amid distractors, $k=6$, 12 principals/point:
\begin{center}\small
\begin{tabular}{lrrrrrr}
\toprule
updates $N$ & 2 & 4 & 6 & 8 & 10 & 12 \\
\midrule
RAG            & 100 & 100 & 100 & 67 & 58 & \textbf{50} \\
\textbf{Tenet} & 100 & 100 & 100 & \textbf{100} & \textbf{100} & \textbf{100} \\
\bottomrule
\end{tabular}
\end{center}
RAG degrades monotonically once $N>k$ ($-50$\,pp); Tenet is flat at $100\%$. Supersession
keeps exactly one current value regardless of churn---the property a \emph{belief state} has
and a document index cannot. \textbf{The curves are identical under a \texttt{gpt-4o-mini} and a \texttt{gpt-4o} reader}: the failure is structural (once $N>k$ the latest value is not reliably retrieved), so a stronger reader cannot rescue RAG.

\subsection{Ablations}
\label{sec:abl}
\textbf{Belief--evidence consistency.} Removing the rule drops Tenet to $55\%$ current-value
accuracy with a $45\%$ stale-leak; adding it restores $100\%$ / $0\%$ leak---the single most
important mechanism. \textbf{Surprise-gating} discards $15\%$ of observations as redundant
with no accuracy change, yielding a bounded store where RAG grows unboundedly.

\subsection{Standardized conflict resolution: MemoryAgentBench FactConsolidation}
On the MemoryAgentBench conflict-resolution benchmark~\cite{mab} (SubEM metric and official reader
prompt verbatim; all 800 questions; Wilson 95\% CIs), ingestion-time supersession with
\textbf{fully deterministic, zero-LLM keys} and a deliberately weak \textbf{local 7B
backbone} scores, pooled over four haystack lengths (6K--262K):
\begin{center}\small
\begin{tabular}{lrrr}
\toprule
pooled & naive-RAG (same reader) & \textbf{Tenet} & published SOTA (mini / gpt-4o)~\cite{freshness} \\
\midrule
FC-SH ($n{=}400$) & 47.8 & \textbf{86.5} [82.8, 89.5] & 78.0 / 94.8 \\
FC-MH ($n{=}400$) & 4.5  & \textbf{30.2} [26.0, 34.9] & 30.2 / 51.5 \\
\bottomrule
\end{tabular}
\end{center}
Single-hop \textbf{exceeds the published gpt-4o-mini-tier SOTA (78.0; our CI excludes it) on
a weaker backbone}; every system in the original benchmark table scores $\le$60 (Zep 7,
Mem0 18, MemGPT 28). Multi-hop exactly ties the mini-tier SOTA, where the original table's
best is $\le$7. Accuracy barely degrades with haystack length (SH $89\!\to\!81$ from
6K$\to$262K): the store is conflict-resolved at ingestion---the property assembly-time
aggregation must re-derive at every read. Multi-hop still degrades with length
($42\!\to\!20$); reported honestly.

To remove the backbone confound entirely, we reimplemented four published memory
mechanisms as arms of the \emph{same} harness (same 7B reader, embedder, SubEM, prompt;
6K+32K cells, $n{=}200$/axis): CAR~[Freshness 2026] 87.5~SH~/~33.0~MH, Mem0-style
consolidation 81.0~/~12.0, HippoRAG-v2-style OpenIE+PPR graph 66.0~/~9.0, MemAgent-style
overwrite memory 44.0~/~16.0 ($n{=}25$). \textbf{Tenet leads every arm on both axes
(90.0~/~36.0 on the same cells)}---ingestion-time supersession matches or beats the best
assembly-time aggregation while doing the consistency work once at write time.
Beyond reimplementations, one released framework run fully \emph{black-box} through its
own ingest-and-retrieve pipeline (ReMe, reme-ai 0.4.1.1; identical Qwen reader/judge for
all arms): on LongMemEval\_S full haystacks ($n{=}100$) \textbf{Tenet 67.0\%
[57.3, 75.4] vs ReMe 34.0\% [25.5, 43.7]} (McNemar $p\approx2\times10^{-6}$; matched
RAG 64.0, blind 0.0), Tenet ahead on every question type.

\subsection{Accurate retrieval at zero ingestion cost: MemoryAgentBench AR}
On MAB's other core competency ($\sim$2{,}000 questions, 197K--534K-token contexts;
official per-benchmark metrics: SubEM for RULER-QA, the official LLM-judge for
LongMemEval(S*), choice accuracy for EventQA; matched gpt-4o-mini reader), Tenet with
\textbf{zero-LLM ingestion} (date-aware structured chunks + embeddings only) averages
\textbf{59.3} --- second only to HippoRAG-v2 ~\cite{hipporag2} (65.1), which runs LLM
OpenIE over every context token at ingestion, and 20+ points above Mem0 (32.6),
Zep (37.5) and MemGPT. Per sub-benchmark: EventQA \textbf{70.7} ($n{=}1{,}500$; Wilson CI
$[68.3, 72.9]$ excludes HippoRAG-v2's 67.6), RULER SH-QA 75.0 (parity with 76),
LME(S*) 46.3 vs 50.7, and RULER MH-QA 45.0 vs 66 --- the honest loss:
Personalized-PageRank graph traversal is genuinely stronger at multi-hop chaining over
narrative text. Together with \S4.5, Tenet leads or ties the published field on two of
MAB's four competencies while being the only published memory framework in this comparison
whose ingestion never calls an LLM.

\subsection{Case study: web-agent trajectory memory (LongMemEval-V2)}
Adapting Tenet's ingestion to LME-V2's web-agent trajectory haystacks~[Wu 2026] (DOM
states, actions; up to 115M tokens), three LLM-free changes---structure-aware chunking,
query cleaning, and Qwen3-Embedding{+}BM25 hybrid retrieval (RRF)---lifted gold-evidence
recall from a naive port's $\sim$12\% to \textbf{59.7\%} @48K budget (63.3\% @72K;
corpus ceiling $\sim$92\%) at $\sim$0.25\,s query latency. End-to-end accuracy is
reader-gated, not retrieval-gated---and the gate is the \emph{reasoning mode}, not
precision: 4-bit local 45.1\% [39.5, 50.8], 8-bit 40.3\%, full-precision hosted
\emph{without} extended thinking 40.7\% (over-abstains), all near the same
$P(\text{correct}\,|\,\text{gold})\approx0.5$--$0.6$ extraction ceiling; the
leaderboard-default extended thinking is what converts retrieval headroom, and lies
outside our compute budget. The substrate transfers; the binding constraint is
measurable and external to the memory.

\section{Limitations}
\textbf{Multi-session synthesis} is the one category where RAG still leads ($42.9$ vs $57.1$).
Belief-anchored expansion lifted it from $28.6$ but does not close it: these questions need
evidence from \emph{several} sessions, and expansion only deepens the sessions the top-$k$
already surfaced---if a required session is not among them, its detail is still missing.
Session-diverse retrieval is the natural next step; elsewhere Tenet is at or above RAG.
\textbf{The frontier is a knob, not free lunch}: parity accuracy costs RAG-equal tokens, and
the $1.6\times$ per-token win is at the efficiency point, which trades ${\sim}5$\,pp of raw
accuracy. \textbf{Evaluation}: $n{=}40$, off-Qwen (gpt-4o / gpt-4o-mini readers, local
embedder), one seed; reader stochasticity is ${\approx}{\pm}5$--$7$\,pp, so the one-shot result
is reported as \emph{parity}, not a win. The shipped system uses Qwen Cloud.

\section{Conclusion}
Treating agent memory as a \textbf{self-consistent belief state} rather than a document index
makes it robust to the way real knowledge behaves: it changes. Tenet stays correct under
knowledge churn where retrieval memory collapses, matches a strong RAG's one-shot accuracy at
equal token budget while offering the best accuracy-per-token we tested, with a light
graph-free substrate and no LLM in the read path---and gets time-travel and principled
forgetting for free. We hope \emph{knowledge churn} becomes a standard axis for
evaluating agent memory.

\begin{thebibliography}{9}\small
\bibitem{mem0} Chhikara et al. \emph{Mem0: Production-Ready AI Agents with Scalable Long-Term Memory}. arXiv:2504.19413, 2025.
\bibitem{longmemeval} Wu et al. \emph{LongMemEval: Benchmarking Chat Assistants on Long-Term Interactive Memory}. arXiv:2410.10813, 2024.
\bibitem{zep} Rasmussen et al. \emph{Zep: A Temporal Knowledge Graph Architecture for Agent Memory}. arXiv:2501.13956, 2025.
\bibitem{memgpt} Packer et al. \emph{MemGPT: Towards LLMs as Operating Systems}. arXiv:2310.08560, 2023.
\bibitem{mab} Hu, Wang, McAuley. \emph{MemoryAgentBench: Evaluating Memory in LLM Agents via Incremental Multi-Turn Interactions}. arXiv:2507.05257, 2025.
\bibitem{freshness} \emph{Don't Ask the LLM to Track Freshness: A Deterministic Recipe for Memory Conflict Resolution}. arXiv:2606.01435, 2026.
\bibitem{memstrata} \emph{Temporal Validity in Retrieval Memory: Eliminating Stale-Fact Errors for AI Agents over Evolving Knowledge}. arXiv:2606.26511, 2026.
\bibitem{engram} \emph{Less Context, More Accuracy: A Bi-Temporal Memory Engine for LLM Agents}. arXiv:2606.09900, 2026.
\bibitem{toki} \emph{TOKI: A Bitemporal Operator Algebra for Contradiction Resolution in LLM-Agent Persistent Memory}. arXiv:2606.06240, 2026.
\bibitem{hipporag2} Guti\'errez et al. \emph{From RAG to Memory: Non-Parametric Continual Learning for Large Language Models}. arXiv:2502.14802, 2025.
\bibitem{qwenlong} Shen et al. \emph{QwenLong-L1.5: Post-Training Recipe for Long-Context Reasoning and Memory Management}. arXiv:2512.12967, 2026.
\bibitem{agemem} Yu et al. \emph{Agentic Memory: Learning Unified Long-Term and Short-Term Memory Management for LLM Agents}. arXiv:2601.01885, ACL 2026.
\bibitem{actmem} Zhang et al. \emph{ActMem: Bridging the Gap Between Memory Retrieval and Reasoning in LLM Agents}. arXiv:2603.00026, 2026.
\bibitem{napmem} \emph{NapMem: Memory as a Structured Action Space for Long-Horizon Agents}. arXiv:2607.05794, 2026. Qwen Large Model Application Team, Alibaba.
\bibitem{memflow} \emph{MemFlow: Training-Free Intent-Routed Memory Orchestration with LLM Validation}. arXiv:2605.03312, 2026.
\bibitem{memreader} \emph{MemReader: Active Extraction for Agent Memory Construction}. arXiv:2604.07877, 2026.
\bibitem{memalpha} \emph{Mem-$\alpha$: Reinforcement-Learned Memory Construction for LLM Agents}. OpenReview, 2026.
\bibitem{memcog} \emph{MemCog: Cognitively-Inspired Active Navigation of Agent Memory}. arXiv:2605.28046, 2026. (Affiliation unverified.)
\bibitem{friston} Friston. \emph{The free-energy principle: a unified brain theory?} Nat.\ Rev.\ Neurosci., 2010.
\end{thebibliography}

\end{document}
