\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=violet,urlcolor=violet,citecolor=violet]{hyperref}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.4ex}{0.7ex}
\titlespacing*{\subsection}{0pt}{1.0ex}{0.5ex}
\setlength{\parskip}{0.35em}
\setlength{\parindent}{0pt}

\newcommand{\tenet}{\textsc{Tenet}}
\newcommand{\ci}[2]{{\footnotesize[#1,\,#2]}}

\title{\textbf{Tenet: Agent Memory as a Self-Consistent Belief State}\\
\large Bi-temporal supersession, belief--evidence consistency, and LLM-free reads\\
for long-horizon agents over changing knowledge}
\author{Anas\\\small \texttt{https://github.com/Nas01010101/tenet}}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent Long-term memory for LLM agents is almost universally implemented as
\emph{retrieval over a growing log of past turns}. We argue this is the wrong abstraction
for an agent that must model a \emph{changing} world, and we characterize the failure as
\textbf{knowledge churn}: as a fact is updated $N$ times over a long interaction, stale
versions crowd the retrieval budget $k$, and once $N>k$ the current value is no longer
reliably retrieved. On a controlled benchmark a strong RAG memory falls from $100\%$ to
$50\%$ current-value accuracy as a fact is updated $2\!\to\!12$ times---under both a
\texttt{gpt-4o-mini} and a \texttt{gpt-4o} reader, so the collapse is structural, not a
reader artifact. We propose \tenet{}, which reframes memory as a \textbf{self-consistent
belief state}: (i) turns are distilled into atomic facts with stable semantic keys;
(ii) a \textbf{bi-temporal} record lets a changed fact \emph{supersede} its predecessor at
ingestion (history kept, time-travel queryable); (iii) a \textbf{belief--evidence
consistency rule} retires raw evidence that echoes a superseded belief; (iv) a
\textbf{surprise-gated write policy} stores only observations the current state does not
already predict; and (v) \textbf{belief-anchored evidence expansion} spends spare context
on raw turns from the sessions the belief state already surfaced. The read path never
calls an LLM. \tenet{} holds $100\%$ on this templated single-attribute churn primitive at
every update level (\S\ref{sec:churn}); on the harsher \emph{paraphrased} multi-attribute
ChurnBench (\S\ref{sec:churnbench}) it was initially \emph{falsified} (worst arm); a
read-time consistency rule plus write-side embedding key-resolution---both defaulted
on---recover it to $100/100/100$ at $U{=}2/8/32$ (half-life $32$), tying delete-outright
Mem0-style for first while keeping LLM-free reads. On
MemoryAgentBench \textsc{FactConsolidation} (ICLR~2026) it scores $86.5\%$ single-hop
\ci{82.8}{89.5}---above the published \texttt{gpt-4o-mini}-tier state of the art
($78.0$) despite using a local 7B backbone---and ties the multi-hop state of the art
($30.2$); on MemoryAgentBench \textsc{Accurate Retrieval} it averages $59.3$, second only
to HippoRAG-v2 ($65.1$) among published systems while being the only one whose ingestion
never calls an LLM, and beats the field on EventQA ($70.7$ vs $67.6$; Wilson CI excludes).
All numbers are reproducible from released code with official metrics and prompts verbatim.
\end{abstract}

\section{Introduction}
An agent that talks to a user for months does not need a transcript; it needs a
\emph{model of the user} that stays true as the user changes. Yet the dominant memory
design---retrieval-augmented generation over stored turns~\cite{mem0,longmemeval}---treats
memory as a document index: embed every turn, retrieve the top-$k$ most similar at query
time, and let the reader sort it out. This works for one-shot recall of a \emph{static}
fact. It fails, quietly, when facts \emph{change}.

We formalize this failure as \textbf{knowledge churn}. A user who moves cities several
times generates many ``I moved to $X$'' turns; all are similar to the query ``where do I
live?'', so the top-$k$ fills with \emph{stale} versions. Once the number of updates $N$
exceeds the budget $k$, the latest value may not be retrieved at all; even when it is, the
reader must infer recency from a pile of contradictory statements. Accuracy collapses
(\S\ref{sec:churn}), and---critically---a stronger reader cannot rescue it, because the
failure occurs before the reader sees anything.

The root problem is the abstraction. A document store has no notion that ``I live in
Boston'' and ``I live in Seattle'' are the \emph{same fact} with a \emph{changed value};
both are just passages. We argue memory should instead be a \textbf{belief state}: a
compact set of \emph{current} beliefs about the world, each with a temporal extent,
updated by observation, kept internally consistent, and queryable across time. This is
the stance world-model and predictive-coding accounts take toward
perception~\cite{friston}; we bring it to agent memory and pay particular attention to
\emph{where} consistency is enforced: at ingestion, so the store never contains a stale
current value, rather than at assembly time, where every read must re-derive it.

\paragraph{Contributions.}
\begin{enumerate}
\item We identify and name \textbf{knowledge churn}, a structural failure mode of
  retrieval-augmented memory, give a controlled benchmark exhibiting it (RAG:
  $100\%\!\to\!50\%$ as a fact is updated $2\!\to\!12\times$; reader-invariant), and show
  a belief-state memory is immune on this templated primitive (\S\ref{sec:churn})---then
  stress-test the claim on a harder paraphrased, multi-attribute variant (ChurnBench,
  \S\ref{sec:churnbench}) that falsifies it and motivates the read-time consistency fix
  (\S\ref{sec:churnfix}).
\item We present \tenet{}: bi-temporal keyed supersession at ingestion, a
  \textbf{belief--evidence consistency rule} that retires stale raw evidence, a
  \textbf{surprise-gated write policy}, salience-decay forgetting, and budget-bounded
  \textbf{belief-anchored evidence expansion}---in a light, graph-free store whose read
  path never calls an LLM (\S\ref{sec:method}).
\item On the standardized \textbf{MemoryAgentBench}~\cite{mab} we evaluate both core
  competencies with official metrics and prompts verbatim, Wilson CIs, and matched-reader
  controls. \textsc{FactConsolidation}: single-hop $86.5$ \ci{82.8}{89.5}, above the
  published mini-tier SOTA ($78.0$~\cite{freshness}, CI excludes) on a \emph{weaker}
  (local 7B) backbone; multi-hop $30.2$, tying the same SOTA, where every system in the
  original benchmark table scores $\le 7$. \textsc{Accurate Retrieval}: average $59.3$,
  second among published systems to HippoRAG-v2's $65.1$~\cite{hipporag2}---which runs
  LLM OpenIE over every context token, where \tenet{} ingests with embeddings
  only---and ahead of the field on EventQA ($70.7$ vs $67.6$, CI excludes)
  (\S\ref{sec:mab}).
\item We reimplement four published memory methods (Mem0-style consolidation, CAR
  aggregation~\cite{freshness}, a HippoRAG-v2-style OpenIE+PPR graph, and
  MemAgent-style~\cite{memagent} overwrite memory) as arms in the \emph{same} harness with
  the \emph{same} backbone, isolating the memory mechanism as the only variable
  (\S\ref{sec:baselines}).
\end{enumerate}

\section{Related work}\label{sec:related}
\textbf{Retrieval memory.} Mem0~\cite{mem0} distills salient facts at write time over a
vector store with entity links; it attaches only a \emph{creation} timestamp and notably
removed its graph variant after finding it $3\times$ slower for a thin gain---evidence we
take seriously in choosing a light vector substrate. LongMemEval~\cite{longmemeval} is the
standard long-horizon benchmark; its V2~\cite{lmev2} adds a latency-aware metric,
signalling a field shift toward accuracy \emph{per cost}, which our per-token results
target.

\textbf{Temporal knowledge graphs.} Zep/Graphiti~\cite{zep} maintains a \emph{bi-temporal}
knowledge graph (valid + transaction time) with edge invalidation---the closest prior
system in temporal semantics. It differs in substrate and mechanism: invalidation is
LLM-mediated at graph-construction time over (subject, predicate, object) edges, reads
traverse a graph, and neither a belief--evidence consistency rule over raw text nor
surprise gating is present. \tenet{} keeps the bi-temporal semantics without the graph,
and its supersession is deterministic given the distilled key.

\textbf{The 2026 bi-temporal convergence.} Concurrently with this work, several systems
adopted bi-temporal supersession: MemStrata~\cite{memstrata} applies a deterministic
(subject, relation, object) supersession rule over a bi-temporal ledger with no LLM in
the read path---and shows \emph{similarity-threshold} supersession leaks stale values
where deterministic keying does not, independently corroborating our keyed design;
Engram~\cite{engram} pairs a bi-temporal knowledge graph with a hybrid
facts-plus-raw-chunks read path (converging on our dual-pool finding);
TOKI~\cite{toki} gives contradiction resolution a formal bitemporal operator algebra. On
conflict-resolution benchmarks, \cite{freshness} shows the \emph{assembly} step dominates:
deterministic $\max(\text{serial})$ aggregation after retrieval sets the published SOTA on
MemoryAgentBench FactConsolidation. \tenet{} differs from all of these in \emph{where}
consistency is enforced---at ingestion (the store never contains a stale current value)
\emph{and} at recall (belief--evidence consistency retires stale raw evidence)---and in
coupling the belief state to a write policy (surprise gating) and a budget-bounded
evidence expansion, evaluated on the knowledge-churn axis none of them report.

\textbf{Memory in the Qwen ecosystem.} Alibaba's 2026 memory line takes the opposite
design point: QwenLong-L1.5~\cite{qwenlong} reaches $76.4$ on LongMemEval by RL-training a
30B reasoner that reads up to 128K tokens per query; AgeMem~\cite{agemem} RL-trains the
memory policy itself; ActMem~\cite{actmem} (Alibaba-co-authored) reaches $75.6$ with
LLM-built causal graphs at ingestion. All three spend heavily---at training time,
ingestion time, or read time. \tenet{} reaches $60.0$ on the same benchmark at the same
reader tier with zero-LLM ingestion and $\sim$2K read tokens per query ($\approx 79\%$ of
QwenLong-L1.5's score at under $2\%$ of its read budget), and its structural wins
(churn, conflict resolution) survive on a 7B backbone.

\textbf{Active memory navigation.} A concurrent 2026 line treats memory access itself as
an agentic, learned process. NapMem~\cite{napmem} exposes a four-level memory pyramid
(raw conversation / typed records / topic tracks / user profile) as a structured action
space and trains a GRPO policy that descends the pyramid and stops once evidence is
sufficient---a trained 9B policy beats an untrained 397B model on memory benchmarks.
QwenLong-L1.5~\cite{qwenlong} additionally plans a navigational path over memory chunks,
lifting LongMemEval by $+15.6$ points. MemFlow~\cite{memflow} is training-free: an LLM
router selects a memory tier per query, and an LLM Validator retries with a heavier tier
on a low-confidence read. MemReader~\cite{memreader} moves the same active-navigation
idea to the write side, extracting memory rather than passively logging it. AgeMem
\cite{agemem} and Mem-$\alpha$~\cite{memalpha} RL-train the read/construction policy
itself; MemCog~\cite{memcog} reports a similar cognitive-navigation frame (affiliation
unverified). All of these make the \emph{read} path agentic: an LLM or an RL policy
decides how deep to go and when to stop. \tenet{} takes the opposite bet on the read
side---it keeps recall LLM-free and gets adaptive depth from an embedding saturation
gate (\S\ref{sec:navigate}) instead of a learned or LLM-mediated stop.

\textbf{OS-style and agentic memory.} MemGPT/Letta~\cite{memgpt} pages memory between a
context ``RAM'' and archival ``disk''; A-Mem~\cite{amem} lets the agent link and evolve
notes. Both are largely append-oriented: fact supersession, consistency between distilled
and raw layers, and forgetting are not first-class operations.

\textbf{What is missing.} No prior system combines (a) ingestion-time bi-temporal keyed
supersession, (b) an explicit belief--evidence consistency rule spanning the distilled and
raw layers, (c) surprise-gated writes, (d) principled forgetting, and (e) belief-anchored
budget-bounded expansion in a graph-free store with an LLM-free read path---nor does any
report the knowledge-churn regime.

\section{Knowledge churn}\label{sec:problem}
Let a fact with key $\kappa$ take values $v_1,\dots,v_N$ over an interaction, each update
producing at least one stored turn. A flat retriever with budget $k$ ranks turns by
query similarity; all $N$ versions are near-equally similar to a query about $\kappa$
(paraphrases of the same statement), so the expected number of \emph{stale} passages in
the top-$k$ grows with $N$, and for $N>k$ the \emph{current} value $v_N$ competes with
$N-1$ distractors for $k$ slots with no ranking signal that prefers it. Two consequences:
(1) retrieval of $v_N$ becomes unreliable; (2) even when retrieved, the reader receives
contradictory values and must infer recency from text alone. Both are properties of the
\emph{store}, not the reader---which predicts the reader-invariance we observe in
\S\ref{sec:churn}. A belief state removes the problem at the source: the store contains
exactly one current value per key, so retrieval of the current value is $O(1)$ in $N$.

\section{Method}\label{sec:method}
\tenet{} stores two layers over one bi-temporal table: a \textbf{belief layer}
$\mathcal{B}$ of distilled, keyed facts and an \textbf{evidence layer} $\mathcal{E}$ of
raw turns. Writes maintain consistency; reads are deterministic lookups plus embedding
search---no LLM call.

\subsection{Distillation into keyed beliefs}
Each turn is distilled by a small LLM into atomic facts, each with a stable semantic key
$\kappa=\textit{subject}{::}\textit{attribute}$ (e.g.\ \texttt{user::residence}), a
salience $s\in[0,1]$, and an event time. The key is what makes supersession reliable:
embedding similarity cannot separate a \emph{restated} fact from a \emph{value-changed}
one (we measure a value change at cosine $0.99$---\emph{higher} than a plain rephrasing of
the same fact at $0.79$, so no similarity threshold separates a changed value from a
restatement), but a shared key can. Where the source text is already structured (as in
benchmark fact streams), keys are computed deterministically from the text with no LLM at
all (\S\ref{sec:fc}).

\subsection{Bi-temporal supersession}
Every memory carries event time (\texttt{valid\_at}, \texttt{invalid\_at}) and transaction
time (\texttt{created\_at}, \texttt{expired\_at}). Storing a fact with key $\kappa$ whose
value differs from the current fact at $\kappa$ \emph{supersedes} it: the old fact's
\texttt{invalid\_at}/\texttt{expired\_at} are set; it leaves the current set but remains
in history. Current recall filters \texttt{expired\_at IS NULL};
$\textit{recall}(\textit{as\_of}=t)$ reconstructs the belief set as of any past $t$
(time-travel). Supersession is thus enforced \emph{at ingestion}: the current store never
contains a stale value, and no per-read aggregation is needed to re-derive freshness.

\subsection{Belief--evidence consistency}\label{sec:bec}
The evidence layer lets the reader answer detail questions, but it is also where stale
values hide: a raw turn ``I moved to Boston'' survives even after the belief
\texttt{user::residence} moves on. We therefore retire from current recall any raw slice
$e$ whose embedding is close to a \emph{superseded} belief:
\begin{equation}
\text{exclude } e \;\Leftrightarrow\; \max_{f\in\mathcal{B}_{\text{expired}}}
\cos\!\big(\phi(e),\phi(f)\big) \;\ge\; \tau_{\text{stale}}, \qquad
\tau_{\text{stale}}=0.80 .
\end{equation}
This single rule turns supersession from a fact-layer nicety into end-to-end correctness:
it takes current-value accuracy from $55\%$ to $100\%$ (\S\ref{sec:ablation}).

\subsection{Surprise-gated writes (predictive coding)}
A world model stores \emph{prediction error}, not everything. On write, a raw observation
$e$ is discarded if the store already predicts it, i.e.\ it is near-identical to an
existing slice:
\begin{equation}
\text{store } e \;\Leftrightarrow\; \max_{e'\in\mathcal{E}}
\cos\!\big(\phi(e),\phi(e')\big) \;<\; g_{\text{surprise}}, \qquad
g_{\text{surprise}}=0.97 .
\end{equation}
This bounds the store and drops redundant repetition with no accuracy loss
(\S\ref{sec:ablation}).

\subsection{Forgetting and dual-pool recall}
Each memory's rank is relevance times a decay factor
$d = 2^{-\Delta t/h}\cdot(1+\beta\log(1+\textit{uses}))\cdot(0.6+0.8s)$ with half-life
$h=14$\,d; a sweep archives current, unpinned memories with $d$ below a threshold; pinned
identity facts never decay. Retrieval is a \textbf{dual pool}---beliefs for consistency,
evidence for verbatim detail---guaranteeing each a share of the budget.

\subsection{Fact dynamics: learning how facts change}\label{sec:dynamics}
A belief state becomes a world model when it can predict its own drift. The
bi-temporal ledger is a training set for exactly this: every superseded fact is an
observed lifetime ($\texttt{valid\_at}\!\to\!\texttt{invalid\_at}$) and every current
fact a right-censored one. Per key \emph{class} $c$ (the attribute part of
$\kappa$), we place a conjugate Gamma prior on an exponential change rate; with
$n_c$ observed supersessions and total exposure $T_c$, the posterior-predictive
survival is Lomax:
\begin{equation}
P(\text{still valid after } \Delta t \mid c) \;=\;
\Big(\tfrac{\beta_0 + T_c}{\beta_0 + T_c + \Delta t}\Big)^{\alpha_0 + n_c},
\end{equation}
so ``residence'' learns a slow hazard and ``mood'' a fast one, per user, in closed
form, with no LLM. A second ledger statistic captures \emph{ripple}: the smoothed
co-supersession probability $P(c' \text{ changes} \mid c \text{ changed})$; when a
correlated key changes, the survival of its neighbors is discounted
($S \to S^{1+\gamma\,\text{bump}}$), so one observation updates beliefs about
unobserved attributes---the defining world-model behavior. Both surface as a
\emph{confidence} on every recalled fact and an $\textit{uncertain\_facts}()$ list
the agent can use to re-verify with the user. Deliberately, confidence does
\textbf{not} demote a fact's rank: a doubted fact is still the best known answer,
and we measured rank-demotion re-creating the churn failure of \S\ref{sec:churn}
(current value pushed behind distractors) before reverting to annotation-only.
A pre-registered follow-up asked whether the same confidence could route \emph{reader
compute} (extractive / cheap / full tiers): it cannot---on 120 questions, no threshold
configuration in an 84-point sweep saved reader tokens within 2pp of the 91.7\%
all-full-reader baseline, because confidence is a \emph{currency} signal orthogonal to
the relevance errors that dominate extractive mistakes (rank-1 belief correct on
currency but wrong on attribute in a third of routed cases). Calibration is sufficient
for caveats and doubt surfacing; not for compute routing. We report both nulls rather
than shipping either.

\paragraph{A trained neural dynamics module.} The closed-form model is memoryless by
family; we also train a compact (276k-param, 1-layer GRU) \emph{neural temporal point
process} over fact-update events, whose Weibull hazard head admits non-constant
per-key hazards and whose auxiliary heads predict the next key to change (learned
ripple) and the next value (contrastively, over embeddings). On a controlled synthetic
corpus with planted periodic, bursty, and cascading structure---provably outside the
closed-form family---the neural model reduces held-out NLL from 2.76 to 1.99
nats/event (paired-bootstrap gain $0.81$, 95\% CI $[0.70, 0.93]$), halves survival
mis-calibration (KS $0.56\!\to\!0.26$), and predicts the next value at $0.997$
recall@5 vs $0.05$ chance, stable over five seeds. On real counterfactual-update
chains the result is deliberately mixed: better calibration and observed-lifetime
likelihood on FactConsolidation, but the censoring-dominated aggregate NLL loses to a
marginal-rate baseline, and LongMemEval yields too few clean chains ($n{=}21$) to
conclude. The exported model is a 1\,MB numpy artifact evaluated in $106\,\mu$s,
opt-in behind an env flag with default behavior unchanged, and passes the
churn-integrity guard.

\subsection{Belief-anchored evidence expansion}\label{sec:expansion}
Compressing a session into a few keyed beliefs wins churn and efficiency but can drop the
fine detail a multi-hop question needs, even when the right session \emph{is} retrieved.
We recover it without reverting to flat retrieval: the top-$k$ dual-pool result names both
the belief state and the sessions it came from. Given spare context budget $B$, \tenet{}
fills it with additional query-relevant raw turns drawn \emph{only from those
already-surfaced sessions}, subject to the same consistency filter (\S\ref{sec:bec}) so no
stale value re-enters. $B$ is set to the baseline RAG budget, so expansion never spends
more context than flat retrieval. The knob traces an accuracy--efficiency frontier: $m=0$
is the efficiency point; $m$ large under budget $B$ is the parity point.

\subsection{Saturation-gated navigation}\label{sec:navigate}
Belief-anchored expansion (\S\ref{sec:expansion}) draws raw evidence from a session
boundary; associative multi-hop recall goes further by re-conditioning the cue on
retrieved evidence and re-scoring the store, but at a caller-fixed depth
$\textit{hops}=N$. A fixed $N$ over-fetches simple queries (the extra hops add
distractors that hurt the reader) and under-fetches genuinely multi-hop ones.
\texttt{navigate()} replaces the fixed depth with an iterative, budget-bounded
hop-deepening loop that reuses belief-anchored expansion at every hop and adopts hop
$d{+}1$ only while its newly-reached evidence is still informative:
\begin{equation}
\text{adopt hop } d{+}1 \;\Longleftrightarrow\;
\max_{e\,\in\,\textit{new}_{d+1}} \textit{score}(e) \;\ge\; \tau_{\text{gain}},
\qquad \tau_{\text{gain}} = 0.15,
\end{equation}
where $\textit{new}_{d+1}$ is the evidence reached at depth $d{+}1$ but not at depth
$d$. When no hop clears the gate, or the hop budget ($\le 4$, mirroring NapMem's
tool-call budget) is exhausted, navigation stops and returns the last adopted pool,
dropping the marginal hop's distractors rather than keeping them. The gate is an
embedding-only marginal-relevance test over scores \tenet{} already computes for
recall, so it adds no LLM call and requires no training: it is the training-free
counterpart to NapMem's learned stop and to MemFlow's LLM-Validator retry, running at
embedding-search latency ($\sim$ms) instead of an LLM round trip.

We validate the mechanism, not a benchmark number. On a deterministic, torch-free
synthetic store (hashed bag-of-words embeddings, no network call), \texttt{navigate()}
(a) reaches a bridge fact invisible to broad top-$k$ retrieval via an associative hop,
and (b) stops early (hop 2) on a saturated single-hop query while continuing to hop 3
on a genuinely multi-hop one, confirming the gate behaves as intended (multi-hop reach,
single-hop early-stop). A subsequent seeded A/B on \textsc{FactConsolidation} 6k cells
at the \texttt{qwen3.7-plus} tier ($n{=}20$ per cell, same reader, prompts, and seed;
retrieval-pool construction the only variable) measured \emph{no benefit}: accuracy
identical to default recall on both cells, and $-5$ to $-10$pp against a fixed-4-hop
arm, all within Wilson CIs. Dumped trajectories confirm the mechanism executed as
designed (pools differed from baseline in $21/21$ rows, adaptive depth $2$--$3$): a
sufficiently strong reader is robust to the larger fixed pool, so adaptive
early-stopping buys no answer precision at this tier. We therefore report navigation
as a mechanism contribution only, with a measured null at the tested tier.

\section{Experimental setup}\label{sec:setup}
\textbf{Benchmarks.} (1) A controlled \textbf{knowledge-churn} benchmark (one fact updated
$N$ times amid distractors); (2) a controlled \textbf{knowledge-update} set for ablations;
(3) \textbf{LongMemEval\_S}~\cite{longmemeval} ($\sim$115K-token histories); (4--5) both
core competencies of \textbf{MemoryAgentBench}~\cite{mab} (ICLR 2026):
\textsc{FactConsolidation} (conflict resolution; all 800 questions) and
\textsc{Accurate Retrieval} ($\sim$2{,}000 questions over 22 contexts of 197K--534K
tokens).

\textbf{Protocol fidelity.} For MemoryAgentBench we use the official metrics and prompts
\emph{verbatim} from the benchmark code: SubEM for FactConsolidation and RULER-QA;
multiple-choice accuracy for EventQA; the official LLM-judge templates for
LongMemEval(S*). Readers are matched to the published tier being compared
(\texttt{gpt-4o-mini} for AR; a deliberately \emph{weaker} local \texttt{qwen2.5:7b} for
FC). Every \tenet{} number is paired with a naive-RAG control sharing the reader,
chunking, and embedder, so the memory mechanism is the only variable.

\textbf{Integrity rules.} API failures are excluded, never scored as wrong (0 exclusions
in FC; exclusions reported where they occur); abstentions are scored wrong; ingestion
caches are keyed by the hash of the exact chunk list; we report Wilson 95\% CIs and dump
every miss for error analysis. All numbers reproduce from single documented commands.

\section{Results}\label{sec:results}

\subsection{Knowledge churn}\label{sec:churn}
One fact updated $N$ times amid distractors, $k=6$, 12 principals per point:

\begin{center}
\begin{tabular}{lcccccc}
\toprule
updates $N$ & 2 & 4 & 6 & 8 & 10 & 12\\
\midrule
RAG & 100 & 100 & 100 & 67 & 58 & \textbf{50}\\
\tenet{} & 100 & 100 & 100 & \textbf{100} & \textbf{100} & \textbf{100}\\
\bottomrule
\end{tabular}
\end{center}

RAG degrades monotonically once $N>k$ ($-50$\,pp); \tenet{} is flat at 100\%. The curves
are \emph{identical} under a \texttt{gpt-4o-mini} and a \texttt{gpt-4o} reader: once
$N>k$ the latest value is not reliably retrieved, so a stronger reader cannot rescue
RAG---the failure is structural, exactly as \S\ref{sec:problem} predicts.

\subsection{LongMemEval\_S: the accuracy--efficiency frontier}
$n{=}40$, $k{=}10$, \texttt{gpt-4o} reader (the reader Mem0 and Zep report against),
local embedder; baselines run under identical settings:

\begin{center}
\begin{tabular}{llrrrr}
\toprule
System & mode & recall@10 & QA acc & reader tok & acc/1k tok\\
\midrule
full-context & --- & --- & $\sim$65\%$^{*}$ & $\sim$124{,}000 & 0.5\\
RAG & top-$k$ turns & 95\% & 57.5\% & 2{,}101 & 27.4\\
\tenet{} & efficiency ($m{=}0$) & \textbf{97.5\%} & 52.5\% & \textbf{1{,}067} & \textbf{49.2}\\
\tenet{} & parity (expansion) & \textbf{97.5\%} & \textbf{57.5\%} & 2{,}083 & 27.6\\
\bottomrule
\end{tabular}
\end{center}

At the efficiency point \tenet{} answers at half the context for the best
accuracy-per-token of any system ($1.6\times$ RAG, $\sim$100$\times$ full-context) at
recall parity. Belief-anchored expansion, capped at RAG's own budget, brings one-shot
accuracy to \emph{parity} ($57.5=57.5$) at fewer tokens. On a \texttt{gpt-4o-mini} reader
the parity point edges ahead (60.0 vs 55.0); the per-token dominance holds across
the \texttt{gpt-4o-mini} and \texttt{gpt-4o} readers we ran
($\approx$1.6$\times$). Given reader stochasticity of $\approx\pm5$--7\,pp at
$n{=}40$, we report one-shot accuracy as parity, not a win. The one category still behind
RAG is multi-session synthesis (42.9 vs 57.1, up from 28.6; \S\ref{sec:limits}).

\paragraph{Reader-generality at the frontier tier.} We subsequently ran the identical
captured task set (fresh $n{=}40$ sample, seed 0, Qwen distiller) through three 2026
frontier readers from independent families under one cross-family judge
(\texttt{qwen3.7-plus}). An initial pass batched 10 questions per CLI call; a methodology
audit flagged this as contaminable, so we \textbf{re-ran two scriptable readers one call
per item} (\texttt{gpt-5.5}, \texttt{gemini-3.5-flash}) and re-judged---these un-batched
numbers are of record. Batching had inflated absolute accuracy by $\sim$2--17\,pp; the
\emph{directional} claims survive the clean measurement: \tenet{} $\ge$ RAG at both
operating points on both readers ($77.5/77.5$ vs $75.0$ GPT-5.5; $75.0/72.5$ vs $70.0$
Gemini), the efficiency point delivers $\approx 1.9\times$ accuracy-per-token at half the
context, and the multi-session weakness still \emph{inverts} un-batched (RAG $50.0$ vs
\tenet{} $62.5$ on both). Per-cell Wilson CIs overlap at $n{=}40$; the evidence is the
directional replication across two independently-isolated reader families.
Artifacts: \texttt{docs/lme\_unbatched\_*.jsonl}.

\subsection{Ablations}\label{sec:ablation}
On the controlled knowledge-update set, removing the belief--evidence consistency rule
(\S\ref{sec:bec}) drops current-value accuracy from $100\%$ to $55\%$ with a $45\%$
stale-leak rate; restoring it recovers $100\%$/0\%. It is the single most important
mechanism. Surprise gating discards $15\%$ of observations as redundant with no accuracy
change, bounding the store where RAG grows without bound.

\subsection{MemoryAgentBench \textsc{FactConsolidation}}\label{sec:fc}
Serial-numbered facts with counterfactual updates; questions require the \emph{current}
value. SubEM and the official reader prompt verbatim; all 800 questions; Wilson 95\% CIs;
0 API-error exclusions. \tenet{}'s ingestion here is \textbf{zero-LLM}: supersession keys
are computed deterministically from the fact text, so ingestion costs embeddings only, and
the reader is a local \texttt{qwen2.5:7b}---\emph{below} the published mini tier.

\begin{center}
\small
\begin{tabular}{lrrrr}
\toprule
cell & naive-RAG & \tenet{} & SOTA mini$^{1}$ & SOTA gpt-4o$^{1}$\\
\midrule
SH 6K & 36.0 & \textbf{89.0} \ci{81.4}{93.7} & 71 & 99\\
SH 32K & 50.0 & \textbf{91.0} \ci{83.8}{95.2} & 78 & 92\\
SH 64K & 52.0 & \textbf{85.0} \ci{76.7}{90.7} & 81 & 95\\
SH 262K & 53.0 & \textbf{81.0} \ci{72.2}{87.5} & 82 & 93\\
\textbf{SH pooled} & 47.8 & \textbf{86.5} \ci{82.8}{89.5} & 78.0 & 94.8\\
\midrule
MH 6K & 5.0 & \textbf{42.0} \ci{32.8}{51.8} & 34 & 57\\
MH 32K & 3.0 & \textbf{30.0} \ci{21.9}{39.6} & 27 & 50\\
MH 64K & 3.0 & \textbf{29.0} \ci{21.0}{38.5} & 33 & 58\\
MH 262K & 7.0 & \textbf{20.0} \ci{13.3}{28.9} & 27 & 41\\
\textbf{MH pooled} & 4.5 & \textbf{30.2} \ci{26.0}{34.9} & 30.2 & 51.5\\
\bottomrule
\end{tabular}

$^{1}$\cite{freshness}: candidate extraction + $\max(\text{serial})$ aggregation.
\end{center}

Single-hop $86.5$ exceeds the published mini-tier SOTA ($78.0$; our CI excludes it) on a
weaker backbone; every system in the original MAB table scores $\le 60$ (Zep 7, Mem0 18,
MemGPT 28). Multi-hop exactly ties the mini-tier SOTA, where the original table's best is
$\le 7$. Single-hop barely degrades with haystack length ($89\!\to\!81$ from
6K$\to$262K) because the store is conflict-resolved at ingestion---the property
assembly-time aggregation must re-derive at every read. Multi-hop still degrades with
length ($42\!\to\!20$); reported honestly. The identical reader over naive-RAG memory
scores $47.8/4.5$: the mechanism, not the reader, carries the result.

\subsection{MemoryAgentBench \textsc{Accurate Retrieval}}\label{sec:mab}
22 contexts of 197K--534K tokens, four sub-benchmarks, official per-benchmark metrics,
matched \texttt{gpt-4o-mini} reader. \tenet{} ingestion is again zero-LLM (date-aware
structured chunks + embeddings only):

\begin{center}
\small
\begin{tabular}{lrrr}
\toprule
sub-benchmark (official metric) & naive-RAG & \tenet{} & HippoRAG-v2$^{1}$\\
\midrule
RULER SH-QA (SubEM, $n{=}100$) & 74.0 & \textbf{75.0} & 76\\
RULER MH-QA (SubEM, $n{=}100$) & 40.0 & 45.0 & \textbf{66}\\
LongMemEval(S*) (LLM-judge, $n{=}300$) & 46.0 & 46.3 & \textbf{50.7}\\
EventQA (choice acc, $n{=}1500$) & 71.3 & \textbf{70.7} \ci{68.3}{72.9} & 67.6\\
\midrule
\textbf{AR average} & 57.8 & \textbf{59.3} & \textbf{65.1}\\
\bottomrule
\end{tabular}

$^{1}$Strongest AR system in the MAB paper. Other published frameworks: Mem0 32.6,
Zep 37.5, MemGPT $\approx$39.
\end{center}

\tenet{} beats the published field on EventQA ($70.7$ vs $67.6$; CI excludes), reaches
parity on RULER SH-QA, and is second on AR average---behind only HippoRAG-v2, whose
ingestion runs LLM OpenIE over every token of the 197K--534K contexts, where \tenet{}
never calls an LLM at ingestion. RULER MH-QA is the honest loss (45 vs 66):
Personalized-PageRank graph traversal is genuinely stronger at multi-hop chaining over
narrative text. LME(S*) is 4.4 points short after five documented retrieval iterations;
the residual misses are answer-synthesis-bound (both arms identical), not
retrieval-bound.

\subsection{Same-harness reproductions of published methods}\label{sec:baselines}
To remove any backbone confound, we reimplement four published memory mechanisms as arms
of the \emph{same} FactConsolidation harness with the \emph{same} local-7B reader and
embedder: CAR ($\max(\text{serial})$ candidate aggregation~\cite{freshness}), Mem0-style
LLM consolidation (batched ADD/UPDATE against nearest neighbors~\cite{mem0}), a
HippoRAG-v2-style graph (OpenIE triples, synonym edges, Personalized-PageRank blended
with dense scores~\cite{hipporag2}), and MemAgent-style overwrite memory
(question-conditioned rolling rewrite~\cite{memagent}). These are mechanism
reproductions at matched backbone, not the authors' full systems. On the 6K+32K cells
($n{=}200$ pooled per axis; MemAgent subsampled $n{=}25$ at 6K only, as it reads the
full haystack per question):

\begin{center}
\small
\begin{tabular}{lrrrrr}
\toprule
6K+32K pooled & \tenet{} & CAR & Mem0-style & HippoRAG-style & MemAgent-style\\
\midrule
SH & \textbf{90.0} \ci{85.1}{93.4} & 87.5 \ci{82.2}{91.4} & 81.0 & 66.0 & 44.0\\
MH & \textbf{36.0} \ci{29.7}{42.9} & 33.0 \ci{26.9}{39.8} & 12.0 & 9.0 & 16.0\\
\bottomrule
\end{tabular}
\end{center}

\tenet{} leads every reproduced arm on both axes. CAR---the published-SOTA
recipe---is closest (CIs overlap): ingestion-time supersession matches or beats the best
assembly-time aggregation while doing the consistency work once at write time instead of
at every read. Mem0-style consolidation holds up single-hop (81.0) but collapses
multi-hop (12.0); the graph arm is limited by 7B-quality OpenIE (66.0/9.0); overwrite
memory loses information by construction (44.0/16.0).

\subsection{Case study: web-agent trajectory memory (LongMemEval-V2)}\label{sec:lmev2}
LongMemEval-V2~\cite{lmev2} shifts the setting from chat to \emph{web-agent trajectory}
histories (up to 115M tokens; DOM states, actions, URLs) and scores a latency--accuracy
frontier (LAFS) against a strong RAG reference (51.0\% accuracy) using a mandated
Qwen3.5-9B reader. We adapted \tenet{}'s ingestion to this regime as a stress test of
the retrieval substrate. A naive port (bge-small dense over trimmed DOM slices)
retrieved the gold evidence for only $\sim$12\% of the 295 deterministic-metric
questions. Three LLM-free changes rebuilt it: structure-aware chunking that keeps
(action, URL, salient DOM) units together and dedups cross-state DOM; query cleaning;
and hybrid retrieval---Qwen3-Embedding dense fused with BM25 by reciprocal-rank fusion.
Gold-evidence recall rose to \textbf{59.7\%} at a 48K reader budget (63.3\% at 72K)
against a measured corpus ceiling of $\sim$92\%, with query latency $\sim$0.25\,s---the
fast corner of the LAFS frontier. End-to-end accuracy proved \emph{reader-gated}, not
retrieval-gated---and the gate is the \emph{reasoning mode}, not precision: a 4-bit
local reader scores $45.1\%$ \ci{39.5}{50.8}, an 8-bit one $40.3\%$, and the
full-precision hosted reader \emph{without} extended thinking $40.7\%$ (it
over-abstains), all pinned near the same
$P(\text{correct}\,|\,\text{gold retrieved})\approx0.5$--$0.6$ extraction ceiling.
The leaderboard's default---extended thinking, $\sim$8K reasoning tokens per
question---is what converts retrieval headroom to accuracy, and lies outside our
compute budget. The
case study supports the paper's central claim in a second modality: the memory
substrate---LLM-free ingestion and sub-second reads---transfers, and where the system
falls short the binding constraint is measurable and external to the memory.

\subsection{ChurnBench: a parametric stress test that falsifies our own churn claim}\label{sec:churnbench}
\S\ref{sec:churn}'s churn result uses one templated fact, updated up to 12 times amid
distractors---a regime pre-registered to structurally favor supersession as soon as
updates exceed $k$. ChurnBench generalizes it into a harness: a parametric generator
sweeps updates-per-fact $U\in\{2,4,8,16,32\}$ over five keyed attributes
(residence/job\_title/car/phone/gym), each updated via \emph{paraphrased}, first-person
conversational statements (not templated serial-numbered lines) interleaved with
distractor sessions; SubEM scoring; real \texttt{Tenet.ingest\_session} (no hand-tuned
keys). Four arms at matched backbone (\texttt{qwen3.7-plus} reader,
\texttt{bge-small} embedder): \tenet{}, RAG, and Mem0-/HippoRAG-v2-style reused verbatim
from the \S\ref{sec:baselines} reproduction harness. Pre-registered gate: \tenet{}
churn half-life (largest swept $U$ with accuracy $\geq90\%$) $\geq2\times$ the best
baseline, CI-separated at $U{=}8$; $n{=}50$ questions/point, Wilson 95\% CIs.

\begin{center}
\begin{tabular}{lccccc}
\toprule
$U$ & 2 & 4 & 8 & 16 & 32\\
\midrule
RAG & 96.0 & 100.0 & 100.0 & 78.0 & 30.0\\
HippoRAG-v2-style & 100.0 & 100.0 & 100.0 & 78.0 & 30.0\\
Mem0-style & 90.0 & 98.0 & 100.0 & 100.0 & \textbf{100.0}\\
\tenet{} & 62.0 & 44.0 & 44.0 & 44.0 & 46.0\\
\bottomrule
\end{tabular}
\end{center}

\textbf{Result: falsified, not a partial miss.} Churn half-life: \tenet{}${<}2$, RAG
$8$, HippoRAG-v2-style $8$, Mem0-style $32$. \tenet{} is the \emph{worst} of four arms at
every tested $U$, including $U{=}2$ (its 95\% CI excludes every baseline at $U{\in}\{2,4,8\}$)---the
opposite of \S\ref{sec:churn}'s result and of the pre-registered hypothesis. Diagnosing
a failing case by inspecting \texttt{recall()} directly shows the correct current fact
\emph{is} ranked first, but two or three raw turns for already-superseded values are
still in the $k{=}10$ pool alongside it: the write-time stale-echo filter
(\S\ref{sec:bec}, tuned for near-verbatim echoes) does not reliably retire a raw turn
phrased differently from its distilled paraphrase (e.g.\ ``I got promoted, I'm now a
principal analyst.'' vs.\ ``The user's job title is principal analyst.''---cosine falls
under the filter's threshold). The reader then sees several structurally-identical,
unordered conflicting statements with no recency cue and often answers from a stale one.
Mem0-style---which \emph{deletes} superseded memories outright rather than retiring them
into a still-recallable raw pool---has nothing to leak from and stays flat through
$U{=}32$; naive RAG and HippoRAG-v2-style, with no consolidation at all, collapse
$100\!\to\!30$ from $U{=}8\!\to\!32$ exactly as \S\ref{sec:churn} predicts. Diagnosed in
this work; the fix and its measured effect follow immediately below (full curve,
misses, and reproduction command in \texttt{docs/BENCHMARK.md} \S9).

\subsection{The fix: read-time belief-evidence consistency + currency-structured context}\label{sec:churnfix}
The diagnosis names two gaps: \S\ref{sec:bec}'s \texttt{\_STALE\_ECHO} filter is a
\emph{global} cosine bar tuned for near-verbatim echoes, and the reader prompt gives
no recency cue over an unordered pool. Two additive, LLM-free, read-time fixes---the
store and every ChurnBench ingestion cache are untouched, so no re-ingestion is
needed. \textbf{(1) Read-time belief-evidence consistency}: narrower and more
sensitive than the global filter, a raw slice is dropped only if it is close to a
superseded fact whose \emph{key already has a current fact in the top-$k$ pool}---so
the threshold can drop below the global filter's without opening a cross-key
false-positive surface. Current-fact ranking is byte-identical on or off; only
raw-slice pool membership changes. Threshold chosen from a sweep on real $U{=}2$ data
(ground truth via exact substring match against the deterministic attribute pools):
$0.70$ gives 100\% recall on genuinely-stale echoes at a 7\% false-positive rate on
legitimately-current ones (0.60: 81\% FPR; 0.80, the original global threshold: only
20.9\% recall---why it missed this case). \textbf{(2) Currency-structured reader
context}: the tenet arm's context is split into ``Current beliefs:'' (distilled,
dated) then ``Supporting raw context:'' (raw turns, dated) instead of one flat list---product-faithful,
since the store already knows \texttt{is\_current}/\texttt{valid\_at} (the agent
surface dates facts the same way); other arms have no such metadata, so their prompts
are untouched by design.

\begin{center}
\begin{tabular}{lcccc}
\toprule
$U$ & tenet-baseline & tenet+1 & tenet+2 & \textbf{tenet+1+2}\\
\midrule
2  & 60.0 & 90.0 & 82.0 & \textbf{98.0}\\
8  & 42.0 & 84.0 & 82.0 & \textbf{92.0}\\
32 & 36.0 & 80.0 & 70.0 & \textbf{82.0}\\
\bottomrule
\end{tabular}
\end{center}

Same cached stores across all four arms (only recall-time config differs), live
\texttt{qwen3.7-plus} reader, $n{=}50$/point, Wilson 95\% CIs in
\texttt{docs/BENCHMARK.md} \S9.1. \textbf{Result: partial, pre-registered gate.}
tenet+1+2 clears $\geq90\%$ at both $U{=}2$ (98.0) and $U{=}8$ (92.0) but falls short
of Mem0-style's flat 100 at $U{=}32$ (82.0)---the gate's explicit ``ship the fix,
report half-life honestly'' branch, not a full pass. \textbf{Churn half-life:
tenet+1+2 $=8$}, up from $<2$, now tied with RAG and HippoRAG-v2-style, still behind
Mem0-style's 32 (immune by construction---it deletes superseded memories outright).
All regression gates pass with the fix defaulted on (7 deterministic suites;
\S\ref{sec:churn}'s $U{=}8$ stays 100\%; FactConsolidation sh\_6k is provably
invariant---that arm never stores raw slices), so \texttt{recall()}'s
\texttt{consistency\_threshold} now defaults to $0.70$ in shipped code.

\paragraph{Full recovery via write-side key resolution.} The read-time fix attacks stale
raw evidence \emph{after} it is stored; a complementary write-side fix removes it at the
source. The embedding key-resolution of \S\ref{sec:limits} (a conversational update
supersedes a same-subject belief whose attribute-key embeds within $0.78$ and whose text
clears the $0.66$ floor) makes paraphrased updates actually collide at ingestion, so far
fewer stale versions ever enter the raw pool. With both fixes on (the shipped default), a
fresh four-arm run recovers \textsc{ChurnBench} to $100/100/100$ at $U{\in}\{2,8,32\}$
(half-life $32$, $n{=}30$, Wilson CIs in \texttt{docs/BENCHMARK.md}~\S14): \tenet{} now
\textbf{ties delete-outright Mem0-style for first} and dominates the retrieval baselines
(RAG/HippoRAG-v2-style collapse to $30$ at $U{=}32$)---while keeping \textbf{LLM-free
reads}, where Mem0-style pays an LLM add/update per write. The pre-registered falsification
is thus fully closed, and by a mechanism the belief-state design predicts rather than a
benchmark-specific patch.

\subsection{Closing the write-path dependency: a local distiller}\label{sec:localdistill}
Every result above distills with Qwen Cloud (\texttt{qwen3.7-plus}); the write path---turning
a message into keyed JSON facts---is \tenet{}'s only remaining LLM dependency (reads are
already LLM-free, \S\ref{sec:setup}). We probe whether a small model, LoRA-tuned and served
fully offline, closes it: can a 1.5B model reproduce the reference distiller's supersession
behavior with zero cloud calls?

\textbf{Method.} LoRA SFT (bf16, rank 16, 2 epochs, $\sim$490 synthetic message$\to$facts
pairs, assistant-only loss) of Qwen2.5-0.5B/1.5B-Instruct, labels from the cloud reference
with keys \emph{force-canonicalized} per known attribute (one key regardless of phrasing).
An early candidate's fabrication rate of $1.0$ traced to a data-balance bug---only 6\% of
training examples had an empty target---not a capacity limit; rebalancing to $\sim$22\%
empty-target examples (free to generate, no cloud labeling) dropped it to $0.08$--$0.17$ with
F1/key-consistency unchanged. Evaluated on a \emph{decontaminated} held-out set: novel values
and phrasings, 0/66 message overlap with training data (an earlier, contaminated screen
inflated key-consistency by $\sim$0.17 and hid the finding below entirely).

\begin{center}
\begin{tabular}{lccccc}
\toprule
candidate & F1 & fabrication & key-consist. & churn (superseded/6)\\
\midrule
qwen2.5-0.5b-instruct (untuned) & 0.295 & 0.333 & 0.30 & 0/6\\
qwen2.5-1.5b-instruct (untuned) & 0.405 & 1.0 & 0.40 & 5/6\\
tenet-distiller-0.5b-v2 (LoRA) & 0.867 & 0.167 & 0.75 & 3/6\\
tenet-distiller-1.5b-v2 (LoRA) & 0.652 & \textbf{0.0} & \textbf{0.775} & \textbf{6/6}\\
\midrule
qwen3.7-plus (cloud reference)\textsuperscript{*} & 1.0 & 0.0 & 0.707 & --\\
\bottomrule
\end{tabular}
\end{center}
{\footnotesize\textsuperscript{*}measured on the contaminated screen, not the decontaminated
set---reported for context, not as an apples-to-apples comparison.}

\textbf{Finding: key-consistency, not F1, is the load-bearing axis for supersession.} The
untuned baselines cannot supersede reliably (0/6, 5/6) despite non-trivial F1; the 0.5B LoRA
variant has \emph{higher} F1 than the 1.5B one (0.867 vs 0.652) but lower key-consistency
(0.75 vs 0.775) and only 3/6 clean-churn supersessions---a 0.5B-specific
out-of-distribution consistency gap the contaminated screen did not surface. The 1.5B LoRA
model reproduces the reference's supersession behavior fully offline (6/6, zero fabrication)
and its canonical-key training \emph{exceeds} the cloud reference's own key-consistency
(0.775 vs 0.707), because forcing one key per attribute at training time is a constraint ad
hoc cloud prompting does not get for free. It passes an end-to-end supersession gate
(move/manager-change) the untuned baselines fail, and ships as an opt-in
\texttt{LLM\_PROVIDER=ollama} path, making the full learn$\to$supersede$\to$doubt loop run
with zero cloud calls. We report this as a probe, not a production claim: the eval is
$n{=}26$ messages / 8 paraphrase groups, deterministic point estimates without confidence
intervals---directionally strong enough to ship as opt-in, wide-$N$ validation is future
work. Full tables and the LoRA training pipeline: \texttt{docs/BENCHMARK.md} \S10,
\texttt{scripts/distiller\_lora/}.

\section{Discussion}
\textbf{Where the belief state wins.} Both MAB competencies reward what supersession
provides: FactConsolidation directly (the store contains exactly one current value; no
per-read freshness inference), EventQA indirectly (date-aware structured chunks preserve
event order that flat chunking destroys). The churn benchmark isolates the mechanism:
immunity is by construction, not tuning.

\textbf{Where it loses.} Multi-hop chaining over narrative text (RULER MH-QA) rewards
explicit graph traversal; a keyed belief state has no edges to walk. Multi-session
synthesis needs evidence from several sessions at once; expansion only deepens sessions
already surfaced. Both are retrieval-topology limits, not consistency limits---graph-free
was a deliberate trade (Mem0's own ablation found graphs $3\times$ slower for thin gains),
and the gap quantifies what that trade costs.

\textbf{Cost asymmetry.} Among published systems above 50 on AR, \tenet{} is the only one
with LLM-free ingestion \emph{and} LLM-free reads; HippoRAG-v2 spends LLM calls per
ingested token, QwenLong-L1.5 spends 128K read tokens per query, ActMem builds LLM causal
graphs at ingestion. At equal accuracy this asymmetry compounds over an agent's lifetime:
ingestion happens once per fact, reads happen forever, and \tenet{} prices both at
embedding cost.

\textbf{Ingestion-time consistency makes read-time conflict resolution redundant.} We
tested this directly: adding CAR-style read-time \texttt{max(serial)} aggregation
(\cite{freshness}'s trick, the current FactConsolidation SOTA mechanism) as an optional
reader over \tenet{}'s pool changed nothing---FC multi-hop $15.0\to15.0$, LoCoMo
$29.0\to28.0$ ($n{=}20/100$). Because supersession already keeps exactly one current value
per key, there is rarely a duplicate-key group left for an aggregator to collapse; the work
CAR does at every read, \tenet{} does once at write. This is the belief-state thesis stated
as a null result: the aggregation step that other systems need is redundant when the store
is self-consistent by construction. (Retraction is the converse---an explicit ``forget
$X$'' that deletes without replacement is semantically distinct from value supersession; a
tombstone op is implemented but stays opt-in, as removing a fact entirely hurt a
multiple-choice benchmark that needs the retracted context to recognize the retraction.)

\section{Limitations}\label{sec:limits}
\textbf{Verbatim-recall benchmarks favor raw storage.} On LoCoMo-10 ($n{=}500$
stratified, qwen-judged, same reader/judge both arms) naive RAG beats \tenet{}
$38.8$ vs $33.8$ (McNemar $p{=}0.031$), concentrated in temporal and single-hop
questions whose gold answers reward the \emph{exact wording} of a raw turn:
distillation paraphrases that wording away. LoCoMo stresses verbatim multi-session
recall, not knowledge churn; we report the loss plainly (\texttt{BENCHMARK.md}~\S12).
A related audit finding: the published LoCoMo audit's 156 error entries correct
\emph{evidence citations}, not gold answers---an ``audit-corrected'' key changes QA
accuracy by exactly $0.0$\,pp by construction.
\textbf{Natural-language updates under-fire keyed supersession.} On PersonaMem-v2
(NapMem's benchmark; 4-way MC, $n{=}485$, deterministic scoring) a blind no-memory
control scores $34.4\%$ vs both memory arms at $\sim$50.5\%---so memory is load-bearing
and the comparison is valid---but \tenet{} and RAG \emph{tie} ($50.5$ vs $50.9$,
McNemar $p{=}0.92$), and on the retraction subset we predicted \tenet{} would win, RAG
leads if anything ($74.2$ vs $67.7$, ns). Cause, measured: only $3.8\%$ of turns trigger
keyed supersession because conversational ``I've switched to $X$'' rarely produces a
distilled key that collides with the prior belief---the paraphrase gap again. \emph{We
then fixed this}: an LLM-free embedding key-resolution (a new fact supersedes a
same-subject current fact whose attribute-key embedding is cosine $\ge 0.78$ and
value-compatible, with the fact-text similarity floor raised to $0.66$ to reject
shared-salient-word coincidences) plus a distiller constrained away from vague umbrella
keys raises true-fire on an expanded labeled set (including adversarial shared-word
negatives) from $40\%$ to $80\%$ at $0\%$ false-fire, lifts real PersonaMem firing
$3.8\%\!\to\!\sim$7\% and overall accuracy $50.5\!\to\!53.4$, and passes every regression
gate (default on, hardened). It does \emph{not} fully address explicit retractions
(``forget $X$''), which are deletions, not value replacements---a tombstone mechanism is
the separate next step. This bounds our churn claim: it holds when updates carry a stable
attribute (\S\ref{sec:churn}, \S\ref{sec:fc}), and now degrades gracefully rather than
sharply on free-form preference drift. (We repurposed PersonaMem's
full-context task as retrieval, so absolute scores are not vendor-comparable;
\texttt{BENCHMARK.md}~\S13.)
\textbf{Multi-session synthesis} still trails RAG (42.9 vs 57.1) under
\texttt{gpt-4o}-tier readers---though this inverts under 2026 frontier readers
(\S: reader-generality); session-diverse retrieval is the natural next step. \textbf{Multi-hop chaining} trails graph memory (45
vs 66). \textbf{The frontier is a knob, not a free lunch}: parity accuracy costs
RAG-equal tokens; the $1.6\times$ per-token win trades $\sim$5\,pp of one-shot accuracy.
\textbf{Evaluation scale}: LongMemEval numbers are $n{=}40$, one seed, off-Qwen readers
(reader noise $\approx\pm5$--7\,pp; hence ``parity'', not a win); MAB numbers are
full-scale (800 and $\sim$2{,}000 questions) with CIs. \textbf{Distillation quality}: in
free-form conversation, keys come from a small LLM distiller; a missed key match degrades
supersession to ordinary storage (fail-soft, but the churn guarantee is only as good as
the keying). \textbf{Scope}: single-user assistant memory; multi-writer stores and
adversarial updates are untested. \textbf{Stale raw-turn leakage under natural
conversational churn} (\S\ref{sec:churnbench}, measured, not hypothetical): with several
attributes each updated many times via paraphrased statements, already-superseded raw
turns can survive the write-time stale-echo filter and outvote the correct current fact
in the reader's context. \textbf{Partially fixed} (\S\ref{sec:churnfix}): read-time
belief-evidence consistency plus a currency-structured reader context raise churn
half-life from $<2$ to $8$ (both regression-gate-clean, defaulted on), but the combined
fix still falls short of Mem0-style's write-time-consolidation immunity at $U{=}32$ ---
the residual gap, not the whole failure mode, is what remains open.

\section{Conclusion}
Treating agent memory as a self-consistent belief state rather than a document index
makes it robust to the way real knowledge behaves: it changes. \tenet{} stays correct
under knowledge churn where retrieval memory structurally collapses, exceeds the
published single-hop conflict-resolution SOTA at its reader tier on a weaker backbone,
and is the only published-competitive system on MemoryAgentBench whose ingestion and
reads are both LLM-free. The belief-state view also yields time-travel and principled
forgetting for free. The LLM-free read path is also cheap at scale: with a resident
embedding matrix (built once, appended incrementally, matmul at query time) recall stays
$\sim$11\,ms at $10^5$ stored facts and extrapolates to $\sim$24\,ms at $10^6$ (versus
$>1$\,s and $\sim$10\,s for a re-load-per-query implementation), so RAM---not
latency---is the eventual bound (\texttt{docs/SCALE.md}). We hope knowledge churn becomes
a standard axis for evaluating agent memory.

\paragraph{Reproducibility.} All code, harnesses, official-prompt copies, cached
ingestion artifacts, and per-question dumps: \texttt{github.com/Nas01010101/tenet}
(\texttt{docs/BENCHMARK.md} lists one command per reported number).

\begin{thebibliography}{99}\small
\bibitem{mem0} P.~Chhikara et al. Mem0: Building Production-Ready AI Agents with Scalable
Long-Term Memory. arXiv:2504.19413, 2025.
\bibitem{longmemeval} D.~Wu et al. LongMemEval: Benchmarking Chat Assistants on Long-Term
Interactive Memory. arXiv:2410.10813, 2024.
\bibitem{lmev2} D.~Wu et al. LongMemEval-V2: Evaluating Long-Term Agent Memory Toward
Experienced Colleagues. arXiv:2605.12493, 2026.
\bibitem{zep} P.~Rasmussen et al. Zep: A Temporal Knowledge Graph Architecture for Agent
Memory. arXiv:2501.13956, 2025.
\bibitem{memgpt} C.~Packer et al. MemGPT: Towards LLMs as Operating Systems.
arXiv:2310.08560, 2023.
\bibitem{amem} W.~Xu et al. A-MEM: Agentic Memory for LLM Agents. arXiv:2502.12110, 2025.
\bibitem{mab} Y.~Hu et al. MemoryAgentBench: Evaluating Memory in LLM Agents via
Incremental Multi-Turn Interactions. arXiv:2507.05257, ICLR 2026.
\bibitem{hipporag2} B.~Guti\'errez et al. From RAG to Memory: Non-Parametric Continual
Learning for Large Language Models (HippoRAG~2). arXiv:2502.14802, 2025.
\bibitem{freshness} Don't Ask the LLM to Track Freshness: A Deterministic Recipe for
Memory Conflict Resolution. arXiv:2606.01435, 2026.
\bibitem{memstrata} Temporal Validity in Retrieval Memory: Eliminating Stale-Fact Errors
for AI Agents over Evolving Knowledge. arXiv:2606.26511, 2026.
\bibitem{engram} Less Context, More Accuracy: A Bi-Temporal Memory Engine for LLM Agents.
arXiv:2606.09900, 2026.
\bibitem{toki} TOKI: A Bitemporal Operator Algebra for Contradiction Resolution in
LLM-Agent Persistent Memory. arXiv:2606.06240, 2026.
\bibitem{qwenlong} Shen et al. QwenLong-L1.5: Post-Training Recipe for Long-Context
Reasoning and Memory Management. arXiv:2512.12967, 2026.
\bibitem{agemem} Yu et al. Agentic Memory: Learning Unified Long-Term and Short-Term
Memory Management for LLM Agents. arXiv:2601.01885, ACL 2026.
\bibitem{actmem} Zhang et al. ActMem: Bridging the Gap Between Memory Retrieval and
Reasoning in LLM Agents. arXiv:2603.00026, 2026.
\bibitem{memagent} MemAgent: Reshaping Long-Context LLM with Multi-Conv RL-based Memory
Agent. arXiv:2507.02259, 2025.
\bibitem{napmem} NapMem: Memory as a Structured Action Space for Long-Horizon Agents.
arXiv:2607.05794, 2026. Qwen Large Model Application Team, Alibaba.
\bibitem{memflow} MemFlow: Training-Free Intent-Routed Memory Orchestration with LLM
Validation. arXiv:2605.03312, 2026.
\bibitem{memreader} MemReader: Active Extraction for Agent Memory Construction.
arXiv:2604.07877, 2026.
\bibitem{memalpha} Mem-$\alpha$: Reinforcement-Learned Memory Construction for LLM
Agents. OpenReview, ICLR 2026.
\bibitem{memcog} MemCog: Cognitively-Inspired Active Navigation of Agent Memory.
arXiv:2605.28046, 2026. (Affiliation unverified.)
\bibitem{friston} K.~Friston. The free-energy principle: a unified brain theory?
\emph{Nat.\ Rev.\ Neurosci.}, 2010.
\end{thebibliography}

\appendix
\section{Protocol details}
\textbf{SubEM} normalization and the reader prompt for FactConsolidation are copied
verbatim from the MemoryAgentBench repository. \textbf{LongMemEval(S*)} judging uses the
five official anscheck templates (including the abstention template) verbatim; judge
failures exclude the instance from scoring. \textbf{EventQA} uses a copy-one-candidate
reader with the official choice sets. \textbf{Caching}: ingestion caches are keyed by the
MD5 of the exact chunk list, and a store/chunk length assertion guards against stale-cache
pairing. \textbf{Controls}: naive-RAG shares the embedder, chunker, reader, prompt, and
$k$ with \tenet{} in every table.

\section{Reproduction commands}
FactConsolidation: \texttt{LLM\_PROVIDER=ollama OLLAMA\_MODEL=qwen2.5:7b
EMBED\_PROVIDER=local python scripts/bench\_factcon.py --qpc 100 --tenet-read decompose
--keys heuristic}. Accurate Retrieval: \texttt{LLM\_PROVIDER=openrouter
OPENROUTER\_MODEL=openai/gpt-4o-mini JUDGE\_PROVIDER=openrouter JUDGE\_MODEL=openai/gpt-4o
EMBED\_PROVIDER=local python scripts/bench\_mab\_ar.py --qpc 100 --judge}. Churn:
\texttt{python scripts/bench\_horizon.py --principals 12 --k 6 --updates 2,4,6,8,10,12}.
Frontier: \texttt{python scripts/lme\_recall.py --limit 40 --k 10 --qa --seed 2
[--expand 20]}. Ablations: \texttt{python scripts/bench\_knowledge\_update.py
--principals 4}. Baseline arms: \texttt{python scripts/bench\_baselines.py --arms
car,mem0,hipporag,memagent}.

\end{document}
